---
title: "From Messy Data to Clean Maps"
author: "LaTreese Denson & Alex Davis"
date: "06/16/2021"
output:
  html_document:
    toc_float: yes
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Workshop Goal:
Answer the research question: Does reef size impact species distribution? Provide maps to help readers visualize the results.

## Workshop Objectives:
- Gain a working knowledge of data cleaning steps using the Tidyverse R.
- Perform computations and statistical analyses using various base functions and packages in R.
- Produce publishable content using the ggplot package in R.

**Note:** If you have quick questions about how to use these packages check out the cheat sheets:

[ggplot2 cheat sheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)

[dplyr cheat sheet](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&ved=2ahUKEwjp7svH3pntAhVt0FkKHfR0DxYQFjAAegQIBBAC&url=http%3A%2F%2Fwww.rstudio.com%2Fwp-content%2Fuploads%2F2015%2F02%2Fdata-wrangling-cheatsheet.pdf&usg=AOvVaw2dHobUyR72zhFqfVO45D5A)

... or **Google it**!

# Import Data
Read in the [growth_data](https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv) from Github using `read.csv()` and give it the object name "data" using `=` or `->` as the assignment operator:

```{r}
data = read.csv("https://raw.githubusercontent.com/latreesedenson/SEAinCode/main/growth_data.csv")

data = as.data.frame(data)
```
**Note:** You could also copy and paste the data from the web browser into an excel file, change the format to comma delimited, saved it as a csv file and then read it in using `read.csv()`. This way you would have a copy of the data on your local drive to read from. You would need to replace the weblink with the path to the data on your local computer.

# Clean Data
Make sure the data was read in correctly using the function `head()`. And look at parts of the data using the `$` operator. 
```{r}
head(data)

data$temp_range
```
**ProTip:** Not sure what a function does? Type `?head()` into the console and see what pops up in the help menu. Scroll down to the bottom--there are often examples of how to use the function that you can practice with.

What are the column names? Is everything there that should be there?
```{r}
names(data)
```

What are the dimensions of the dataset, how many rows and columns?
```{r}
dim(data)
```


Let's get some basic summary statistics from our data: minimums, maximums and means.
```{r}
summary(data)
```


You will notice that there were multiple species here and some NAs in the data. Let's work with the first species and remove the bad records.
We will use the pipe operator `%>%` from the magrittr package within the tidyverse to do this. This syntax leads to code that is easier to write and to read. Use the keyboard shortcut: Ctrl + Shift + M (Windows) or Cmd + Shift + M (Mac) to get the pipe operator.

```{r}
library(tidyverse)
Species_A_Clean = data %>% filter(species =="A",!is.na(temp_range))

head(Species_A_Clean)
dim(Species_A_Clean)
summary(Species_A_Clean)
```

Notice that there are no more NAs in the temp_range summary.
```{r}
Species_A_Clean$temp_range
```

# Basic Visualization and Analyses
Let's do a boxplot of our response vs. our explanatory variables. Here we are using the ggplot2 package, but I also provide code for using Base R.

ggplot can get a bit complex depending on what you want to do but the basic components are typically the same. 

You will notice that I have commented out some lines of code using '#'. This allows me to block out lines that I don't want to run, as well as create comments for a better explanation of my code. 

```{r, echo=TRUE}
# ggplot general inputs/arguments (data, aesthetics - x and y) + plot type + etc.

ggplot(Species_A_Clean, aes(x=as.factor(daylength), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("day length")

# To save to a file for viewing later we use the function ggsave()
# The .tiff extension is the type of file, you can also use .jpg
# .tiff files are good for publication graphics
# ggsave("boxplot_daylength.tiff")


# Change the x variable to graph a different explanatory variable
ggplot(Species_A_Clean, aes(x=as.factor(temp_range), y=growth)) + 
  geom_boxplot(fill="slateblue", alpha=0.2) +  xlab("temp")

# ggsave("boxplot_temp.tiff")

# Here is the code to create a boxplot and save it in Base R
# Remove the # to run these 3 lines
# tiff("boxplot.tiff")
# boxplot(growth~temp_range,data=Species_A_Clean,xlab="temp_range", ylab="growth(wt")
# dev.off() 
```
Notice the way we save graphics is different depending on the package we are using. Also `dev.off()` closes all graphics devices so you can start a fresh graph next time.

Create a histogram of the data using the `geom_histogram()` function. Change the `binwidth` to see what happens.
```{r}
ggplot(Species_A_Clean, aes(x = growth)) +
  geom_histogram(binwidth = 1)

# You can remove the '#' in the line below if you want to save this
# ggsave("DataDistribution.tiff") 

# Base R
# hist(Species_A_Clean$growth, breaks=1, col="darkgrey", main = "", xlab="growth") 
```


# Mapping
Let's fit a basic linear model to the data, to determine the impact of day length and temperature  on species growth.

We use the `lm()` function which needs model and the data as it's input. The model is setup using a $\sim$, where $y\sim x$ means that Y is a function of a series of Xs.

```r
lm(y ~ x1 + x2 + x3, Data = Data)

```
Let's run a model with our species data. We tell R that we have factors with multiple levels by using the function `factor()`. If you had a continuous variable you wouldn't need this added piece.
```{r}
SpeciesA.lm = lm(growth ~ factor(daylength)*factor(temp_range), data = Species_A_Clean)
```

To get a summary of the linear model output we use the function `summary()`.
```{r}

sum.mod = summary(SpeciesA.lm) 
# since sum.mod is an object 
# we will need to run the next line to see what is in it

sum.mod

####
# Additional functions you may find useful in the future
####
# coefficients(fit) # model coefficients
# confint(fit, level=0.95) # CIs for model parameters
# fitted(fit) # predicted values
# residuals(fit) # residuals
# anova(fit) # anova table 
####
```
Here you have your important outputs to make a conclusion with (pvalues, coefficients, etc.).

Let's save some of that information to an excel file. Note: you cannot save the entire output to a csv using the lines below. You will need additional steps and functions.
```{r}
# save results to an object
results = sum.mod$coefficients

# make the object into a dataframe so it is delimited
# try it without as.data.frame to see what I mean
out = as.data.frame(results)

# Write to a csv file 
write.csv(out,"ModelCoefficients.csv")  

# You can also write to a text file using write.table() and .txt file name instead of csv

```

Let's plot the diagnostics of the model fit to the data.`plot(linearmodel)` produces 4 main plots; each used to diagnose the fit of the model to the data.
```{r}
# Remember you can save this graphic by uncommenting the line below
# be sure to use dev.off after to reset things

# tiff("ModelDiagnostics.tiff")
# The graphing format: 2 rows 2 columns
par(mfrow=c(2,2)) 

# type dev.off and run the next line without this function to see what happens
plot(SpeciesA.lm)

# close the plotting window and resets things
dev.off()  
```

# Final ProTips
1. Comment your code! ( Use the hashtag - # )
2. R is case-sensitive. For example, "head()" is not the same as "Head()." 
3. Be smart when naming files, folders, etc.. You will be looking for them later. Adding the date or sequential numbers to a folder name is helpful.
4. Debugging is the hardest part. Copy the error you get and google it. The [Stack Overflow](https://stackoverflow.com) community will be your best friend.
5. There are a million ways to skin a cat! What I showed you here is only one way of using R to do the things we want to do. Find what works for you.
